Thesis Title

### 2022-08-08 chapters/introduction.tex
Add Grammarly suggestions - thesis overview and introduction

\begin{figure*}[h!]
\centering
\includegraphics[width=\textwidth]{figures/introduction/Figure 1.pdf}
\caption{\textbf{World Health Organisation Global Priority Pathogens list.} This catalogue includes, besides \textit{Mycobacterium tuberculosis} considered the number one global priority, a list of twelve microorganisms grouped under three priority tiers according to their antimicrobial resistance: critical (\textit{Acinetobacter baumannii}, \textit{Pseudomonas aeruginosa} and \textit{Enterobacteriaceae}), high (\textit{Enterococcus faecium}, \textit{Helicobacter pylori}, \textit{Salmonella} species, \textit{Staphylococcus aureus}, \textit{Campylobacter} species and \textit{Neisseria gonorrhoeae}), and medium (\textit{Streptococcus pneumoniae}, \textit{Haemophilus influenzae} and \textit{Shigella} species). The major objective was to encourage the prioritisation of funding and incentives, align research and development priorities of public health relevance, and garner global coordination in the fight against antimicrobial resistant bacteria. Adapted from \cite{world_health_organization_prioritization_2017}.}
\label{fig:figure1}
\end{figure*}

@@

The past few decades have seen a major revolution in the operation of microbial laboratories, driven by the development of molecular technologies and ways to make these accessible, namely amplification-based \ac{PCR}, matrix-assisted laser desorption/ionisation - time of flight (MALDI-TOF) and \ac{DNA}-microarray-based hybridisation technology. 
These are used in conjunction with traditional techniques such as microscopy, culture, and serology.
Application of these methods differs by suspected infection type: bacterial, viral, fungal or parasitic. 
For the purpose of this dissertation work, we will focus on bacterial (see Section \ref{sssec:_intro_bacterial}) and viral infections (see Section \ref{sssec:_intro_viral}).

@@

The current gold standard for bacterial pathogen identification in diagnostic microbiology laboratories involves the isolation of the pathogen through culture followed by biochemical testing, a multi-step process that can take days to weeks before obtaining results, depending on the fastidiousness of the organism and if it can be cultured \citep{muhamad_rizal_advantages_2020, giuliano_guide_2019, muhamad_rizal_advantages_2020}. 
Although culture allows the identification of a wide variety of organisms, some pathogens can escape routine due to strict metabolic necessities for growth or the requirement for specific biochemical tests needed for their identification. 
Furthermore, results will be obscured if a mixed culture is obtained, particularly if the cultures are obtained from sites with microbiota, such as the gut and the skin, increasing the risk of contamination by normal flora and leading to false results \citep{giuliano_guide_2019}. 
After successful growth in culture, Gram staining and MALDI-TOF mass spectrometry are often used for identification with good accuracy as long as the pathogen is presented in the coexisting database \citep{patel_maldi-tof_2015}. 
An alternate rapid identification method is \ac{PCR} where nucleic acid fragments are detected through specific primers, being highly sensitive and specific, to the point where \ac{PCR} may detect bacteria that are not viable after a patient has been treated for an infection and it is limited to the primer used \citep{scerbo_beyond_2016}. 
Syndromic panels, an extension of \ac{PCR} by using multiple primers (multiplex \ac{PCR}) to simultaneously amplify nucleic acids from multiple targets in a single reaction, tried to address this issue by allowing for the identification of multiple bacteria and other important information such as the detection of antibiotic resistance or virulence genes \citep{giuliano_guide_2019}.

@@

Traditional \ac{PHSS} can have different approaches based on the epidemiology and clinical presentation of the disease and the goals of surveillance. 
In passive surveillance systems, medical professionals in the community and health facilities report cases to the public health agency, which conducts data management and analysis once the data is received and communicates with the responsible entities. 
Globally, the \ac{WHO} as described in the International Health Regulations what is notifiable by all countries, such as severe acute respiratory syndrome (SARS) and viral hemorrhagic fevers (Ebola, Lassa, Marburg), as well as guiding which public health measures should be implemented \citep{world_health_organization_international_2005}. 
Active surveillance aims to detect every case, not relying on a reporting structure, and can have many approaches from sentinel sites or network of sites that capture cases of a given condition, such as respiratory tract infections, within a catchment population \citep{murray_infectious_2017, melo-cristino_estudo_2006}. 
The application of environmental surveillance methods, performed prospectively to detect pathogens prior to the recording of clinical cases or to monitor their abundance in the environment to assess the potential risk of disease, has been proven as a viable alternative, particularly in wastewater \citep{andrews_environmental_2020, mcweeney_demonstration_1894, baker_combined_2011, larsen_tracking_2020}.

@@

\begin{figure*}[h!]
\centering
\includegraphics[width=\textwidth]{figures/introduction/Figure 3.pdf}
\caption{\textbf{Principles of current processing of bacterial pathogens based on whole genome sequencing.} Schematic representation of the workflow for processing samples for bacterial pathogens after the adoption of whole genome sequencing, with an expected timescale that could fit within a single day. The culture steps would be the same as currently used in a routine microbiology laboratory (see Figure \ref{fig:figure2}). Once a likely pathogen is ready for sequencing, \ac{DNA} is extracted, taking as little as 2 hours to prepare the \ac{DNA} for sequencing.
After sequencing, the main processes for yielding information is computational. Automated sequence assembly algorithms are necessary for processing the raw sequence data, from which species, relationship to other isolates of the same species, antimicrobial resistance profile and virulence gene content can be assessed. All the results can also be used for outbreak detection and infectious diseases surveillance. Adapted from \cite{didelot_transforming_2012}}
\label{fig:figure3}
\end{figure*}

The paper describing the \ac{DNA}-sequencing method with chain-terminating inhibitors used in the sequencing of the first microbial genome \citep{sanger_dna_1977}, which earned the late Frederick Sanger his share of the 1980 Nobel Prize in Chemistry alongside Walter Gilbert, was, in 2014, the top fourth in the number of citations with over 60000, highlighting its impact in the field of biological sciences, and by extension medicine \citep{van_noorden_top_2014}. 
Currently, this number has increased to over 84000 according to \ac{PMC}\footnote{\url{https://pubmed.ncbi.nlm.nih.gov/}}\footnote{\url{ https://www.ncbi.nlm.nih.gov/pmc/articles/PMC431765/}}. 
Since its emergence, reductions in cost, technical advances in sequencing technologies, and new computational developments have made genomic sequencing one of the most influential tools in biomedical research, yielding unprecedented insights into microbial evolution and diversity, and the complexity of the genetic variation in both commensal and pathogenic microbes. 
The emerging application of genomic technologies in the clinic to combat infectious diseases is transforming clinical diagnostics and the detection and surveillance of outbreaks.

@@

\begin{figure*}[h!]
\centering
\includegraphics[width=\textwidth]{figures/introduction/Figure 5.png}
\caption{\textbf{The three revolutions in sequencing technology that have transformed the landscape of bacterial genome sequencing.} The first-generation, also known as Sanger sequencers, is represented by the ABI Capillary Sequencer (Applied Biosystems). During the sequencing reaction, at each nucleotide incorporation event, a fluorescently labelled \ac{ddNTP} is incorporated, terminating the elongation of the \ac{DNA} molecule. The resulting electropherogram for sequencing reaction is below, and is read from left to right. The second-generation, also known as high-throughput sequencers, is represented by MiSeq, a 4-channel sequencer, and NextSeq, a 2-channel sequencer (Illumina), both sequencing by synthesis instruments. For both instruments, the loaded flowcell is sequenced in massive parallel reactions, with each nucleotide incorporation emitting a light signal that is captured and latter basecalled into a fastq file, with indication of the confidence of the call, presented below. In a 4-channel instrument each nucleotide has its own marker (A: yellow, T: green, C: red, G: blue) but in a 2-channel instrument only 2 markers exist (A: green plus red, T: green, C: red, G: no marker). These instruments allow the sequencing of both ends of the \ac{DNA} fragment. Lastly, the third-generation, also known as long-read sequencers, is represented by the Pacific Bioscience BS sequencer and Oxford Nanopore MinION sequencer. In the first, immobilised polymerases in a \ac{SMRT} Cell incorporating nucleotides with identifying fluorescent labels. In the latter, a nanopore embedded in a solid-state membrane causes a change in an ionic current across the membrane each time a nucleotide is pushed through the pore. This difference in potential is then used for basecalling. Adapted from \cite{hagemann_overview_2015, loman_twenty_2015,goodwin_coming_2016, wang_nanopore_2021, metzker_sequencing_2010, xu_recent_2020}}
\label{fig:figure5}
\end{figure*}

@@

The release of the first truly high-throughput sequencing platform in the mid-2000s heralded a 50,000-fold drop in the cost of \ac{DNA} sequencing in comparison with the first-generation technologies and led to the denomination of the second generation as next-generation sequencing (NGS) \citep{goodwin_coming_2016}. 
This trend has continued over the next two decades of continued development and improvement, allied to the emergence of benchtop sequencing platforms with a high-throughput sequencing data and turnaround times of days, making it a standard in any microbiology and public health laboratories \citep{loman_twenty_2015}. 
Second-generation sequencing methods can be grouped into two major categories: (1) sequencing by hybridisation and (2) sequencing by synthesis.

@@

The latter is defined by their use of terminator molecules that are similar to those used in the first-generation of sequencing, preventing elongation of the DNA molecule, but unlike the first methods, it is reversible. 
To begin the process, a \ac{DNA} template is primed by a sequence that is complementary to an adapter region, which will initiate polymerase binding to this double-stranded \ac{DNA} region. 
During each cycle, a mixture of all four individually labelled and 3'-blocked \ac{dNTP}s are added. 
After incorporation of a single \ac{dNTP} into each elongating complementary strand, the unbound \ac{dNTP}s are removed and the surface is imaged to identify which \ac{dNTP} was incorporated at each cluster by optical capture. 
The fluorophore and blocking group can then be removed and a new cycle can begin \citep{goodwin_coming_2016}. 
The Illumina systems, which use this technology, accounts for the largest market share for sequencing instruments compared to other platforms\footnote{\url{https://www.forbes.com/companies/illumina/?sh=774358a91aa6}}, allowing paired-end sequencing and having the highest throughput (from 25 million reads for a MiSeq instrument to  1.2 billion reads for a NextSeq instrument\footnote{\url{https://www.illumina.com/systems/sequencing-platforms.html}}), with read lengths ranging from 45 to 300 bases in length with high accuracy, albeit with long running times (4 to 55 hours), rendering this technology a good choice for many sequencing applications where large read length is not required \citep{loman_performance_2012, gupta_chapter_2014, hagemann_overview_2015}.

@@

\ac{ONT} makes use of nanopores in small, portable single-molecule sequencing devices, capable of generating ultra-long sequences in real-time at a relatively low cost. 
Biological nanopores are embedded in solid-state membranes within disposable flow cells which, when a \ac{DNA} strand passes through the pore driven by a motor protein, each nucleotide causes a change in an ionic current across the membrane, which is later base called \citep{hoang_long-reads-based_2022, loman_twenty_2015}. 
This process is free from fluorescence labels and amplification requirements, and after one strand is processed, the pore is available to sequence the next available strand. 
Sequence quality and length depend on the loaded library but are usually much lower than the alternative counterparts, and its throughput is dependent on the number and lifespan of the nanopore within the flowcell, but still much lower than the alternatives. 
Despite this, its portability, fast advances, and continued improvement of the flowcells make this a fast adopted technology for long-read sequencing.

@@

Currently, sequencing technologies are used in a case-by-case approach, with its adoption being much more present in a research setting than in a diagnostic one. 
Sequencing is mostly used after a diagnostic through the identification of the causative agent has already been performed. 
Although substantial advances have been made in reducing response time, most of the current systems do not yet generate enough data fast enough for a truly rapid response for it to be used in the clinical setting \citep{goodwin_coming_2016}. 
High-throughput \ac{DNA} sequencing has found additional new applications in drug discovery and in functional genomics with, for example, SNP-based analysis to identify new drug targets \citep{loman_twenty_2015}.

Although the second-generation \ac{DNA} sequencing methods have shed light on fundamental aspects of microbial ecology and function, they suffer from issues associated with short read length (see \ref{ssec:_intro_2nd_gen_seq}) and cannot reliably reconstruct long repeats because of uncertainties in mapping read, even when paired-end sequencing is used. 
Third-generation sequencing methods (see \ref{ssec:_intro_3rd_gen_seq} The third-generation of \ac{DNA} sequencing) have become increasingly used in microbiology, although their accuracy and low throughput make it difficult to implement in a clinical diagnostic setting.

@@

In the 2009 swine-origin Influenza A H1N1 pandemic, the first complete genome was publicly available on the 25 of April of 2009 (GenBank accession number FJ966079), about a month after records of increased flu activity in Mexico and 10 days after the first confirmed cases in California, United States of America \citep{smith_origins_2009, novel_swine-origin_influenza_a_h1n1_virus_investigation_team_emergence_2009}. 
By the time the pandemic was declared, on 11 of June of 2009, \cite{smith_origins_2009} reported the origins and evolutionary genomics of the pandemic influenza A variant with a collection of 813 complete influenza genome sets, 17 of which belonging to the newly swine influenza viruses (GenBank accessions numbers GQ229259–GQ229378). 
The \ac{MERS} pandemic, declared as such in 2015 \citep{piret_pandemics_2021}, had its first publicly available sequence on 5 of July 2015 (GenBank accession number KT006149)\citep{lu_complete_2015}, with a sequence from a camel, thought to be an intermediate host for the virus, available as early as 7 of March 2016 (GenBank accession number KU740200) \citep{kandeil_complete_2016, al-shomrani_genomic_2020}.

@@

Molecular barcoding approaches can be combined with second-generation high-throughput sequencing to achieve unprecedented depths of coverage in microbial community profiling, being defined as metataxonomics. 
For profiling bacterial species, the most popular approach is 16S \ac{rRNA} gene sequencing, an ~1500 base pair gene that encodes catalytic RNA that is part of the 30S ribosomal subunit. 
Traditionally, the variable regions of the 16S \ac{rRNA} gene (V-regions) are targeted, or ranges thereof (V1-V2, V1-V3, V3-V4, V4, V4-V5, V6-V8, and V7-V9), and are specific to bacterial genus (96\%) and for some, even species (87.5\%), \citep{srinivasan_use_2015, abellan-schneyder_primer_2021}. 
Moreover, dedicated 16S databases that include near full-length sequences for a large number of strains and their taxonomic placements exist, such as RDP\footnote{\url{http://rdp.cme.msu.edu/}}, Greengenes\footnote{\url{https://greengenes.secondgenome.com/}}, silva\footnote{\url{https://www.arb-silva.de/}} and NCBI's 16S ribosomal RNA project\footnote{\url{https://www.ncbi.nlm.nih.gov/refseq/targetedloci/}} \citep{cole_ribosomal_2009, desantis_greengenes_2006, pruesse_silva_2007}. 
The sequence of an unknown strain can be compared with the sequences in these databases, after very closely related sequences are grouped into \ac{OTUs},  an operational definition used to classify groups of closely related individuals. This allows the deduction of probable taxonomy, with the assumption that sequences of $>$95\% identity represent the same genus, whereas sequences of $>$97\% identity represent the same species \citep{schloss_introducing_2005}.

@@

While viruses are an integral part of the microbiota, no universal viral marker genes are available to perform such taxonomic assignments. 
Amplification of whole viral genomes is possible and, in 2015, RNA extracted from whole blood, serum, re-suspended swabs and urine, after targeted amplification of the whole viral genome, proved invaluable in the track of the Ebola virus disease epidemic in West Africa, responsible for >11 thousand deaths, allowing for the characterisation of the infectious agent the determination of its evolutionary rate, signatures of host adaptation, identification and monitoring of diagnostic targets and responses to vaccines and treatments \citep{quick_real-time_2016}.
As an alternative, broad scope viral targeted sequence capture (TSC) panels offer depletion of background nucleic acids and improve the recovery of viral reads by targeting coding sequence from a multitude viral genera, such as VirCapSeq-VERT Capture Panel\footnote{\url{https://sequencing.roche.com/content/dam/rochesequence/worldwide/resources/brochure-vircapseq-vert-capture-panel-SEQ1000117.pdf}} but do not guarantee the full recovery of the viral genome, and can present biases towards certain genera \citep{schuele_assessment_2020, wylie_enhanced_2015}.

@@

Clinical applications of \ac{SMg} derive its roots from the use of microarrays (see \secref{ssec:_intro_current_standards}), where it was successfully applied in in-depth microbiome analysis of different sites in the human body, it was the emergence of second-generation sequencing technology and its high throughput of genomic data at a competitive price that made sequencing of all genomic content, \ac{DNA} and/or RNA, in a clinical sample a viable possibility for diagnostics (see \secref{ssec:_intro_2nd_gen_seq} The second-generation of \ac{DNA} sequencing) \citep{miller_basic_2009, palmer_rapid_2006, chiu_clinical_2019}. The first reported case that demonstrated the utility of \ac{SMg} was in 2014 with the clinical diagnosis of neuroleptospirosis in a 14-year-old immunodeficient and critically ill boy with meningoencephalitis by Wilson et al \cite{wilson_actionable_2014}, prompting appropriate targeted antibiotic treatment and eventual recovery of the patient. In this case, traditional methods, including an invasive brain biopsy, failed to provide answers, until the shotgun sequencing of cerebrospinal fluid identified 475 of 3,063,784 sequence reads (0.016\%) corresponding to leptospira, for which clinical assays were negative due to its very low abundance. Ever since many other reports of successful application of \ac{SMg} in clinical metagenomics have been reported. but all in edge cases where traditional diagnostic methods have failed or as proof-of-concept \citep{couto_critical_2018, vijayvargiya_application_2019, sanabria_shotgun-metagenomics_2020, hirakata_application_2021}.

@@

As stated previously (see \secref{sec:_intro_genomics_approach} and \secref{ssec:_intro_metagenomics}), one of the biggest challenges when dealing with genomic, and in particular metagenomic, data is the lack of golden standards. This is also applicable to the bioinformatic analysis, required due to the amount of data produced by genomic sequencing technologies. This is currently one of the bottlenecks in the deployment of sequencing technology in clinical microbiology as there is no standard in how to deal with the increasing amount of data produced in a fit-for-purpose manner \citep{carrico_primer_2018}.

>>>>

\begin{figure*}[h!]
\centering
\includegraphics[width=\textwidth]{figures/introduction/Figure 1.pdf}
\caption{\textbf{World Health Organisation Global Priority Pathogens list.} This catalogue includes, besides \textit{Mycobacterium tuberculosis} considered the number one global priority, a list of twelve microorganisms grouped under three priority tiers according to their antimicrobial resistance: critical (\textit{Acinetobacter baumannii}, \textit{Pseudomonas aeruginosa} and \textit{Enterobacteriaceae}), high (\textit{Enterococcus faecium}, \textit{Helicobacter pylori}, \textit{Salmonella} species, \textit{Staphylococcus aureus}, \textit{Campylobacter} species and \textit{Neisseria gonorrhoeae}), and medium (\textit{Streptococcus pneumoniae}, \textit{Haemophilus influenzae} and \textit{Shigella} species). The major objective was to encourage the prioritisation of funding and incentives, align research and development priorities of public health relevance, and garner global coordination in the fight against antimicrobial-resistant bacteria. Adapted from \cite{world_health_organization_prioritization_2017}.}
\label{fig:figure1}
\end{figure*}

@@

The past few decades have seen a major revolution in the operation of microbial laboratories, driven by the development of molecular technologies and ways to make these accessible, namely amplification-based \ac{PCR}, matrix-assisted laser desorption/ionisation - time of flight (MALDI-TOF) and \ac{DNA}-microarray-based hybridisation technology. 
These are used in conjunction with traditional techniques such as microscopy, culture, and serology.
The application of these methods differs by suspected infection type: bacterial, viral, fungal or parasitic. 
For the purpose of this dissertation work, we will focus on bacterial (see Section \ref{sssec:_intro_bacterial}) and viral infections (see Section \ref{sssec:_intro_viral}).

@@

The current gold standard for bacterial pathogen identification in diagnostic microbiology laboratories involves the isolation of the pathogen through culture followed by biochemical testing, a multi-step process that can take days to weeks before obtaining results, depending on the fastidiousness of the organism and if it can be cultured \citep{muhamad_rizal_advantages_2020, giuliano_guide_2019, muhamad_rizal_advantages_2020}. 
Although culture allows the identification of a wide variety of organisms, some pathogens can escape routine due to strict metabolic necessities for growth or the requirement for specific biochemical tests needed for their identification. 
Furthermore, results will be obscured if a mixed culture is obtained, mainly if the cultures are obtained from sites with microbiota, such as the gut and the skin, increasing the risk of contamination by normal flora and leading to false results \citep{giuliano_guide_2019}. 
After successful growth in culture, Gram staining and MALDI-TOF mass spectrometry are often used for identification with good accuracy as long as the pathogen is presented in the coexisting database \citep{patel_maldi-tof_2015}. 
An alternate rapid identification method is \ac{PCR} where nucleic acid fragments are detected through specific primers, being highly sensitive and specific, to the point where \ac{PCR} may detect bacteria that are not viable after a patient has been treated for an infection and it is limited to the primer used \citep{scerbo_beyond_2016}. 
Syndromic panels, an extension of \ac{PCR} by using multiple primers (multiplex \ac{PCR}) to simultaneously amplify nucleic acids from multiple targets in a single reaction, tried to address this issue by allowing for the identification of multiple bacteria and other important information such as the detection of antibiotic resistance or virulence genes \citep{giuliano_guide_2019}.

@@

Traditional \ac{PHSS} can have different approaches based on the epidemiology and clinical presentation of the disease and the goals of surveillance. 
In passive surveillance systems, medical professionals in the community and health facilities report cases to the public health agency, which conducts data management and analysis once the data is received and communicates with the responsible entities. 
Globally, the \ac{WHO} as described in the International Health Regulations what is notifiable by all countries, such as severe acute respiratory syndrome (SARS) and viral hemorrhagic fevers (Ebola, Lassa, Marburg), as well as guiding which public health measures should be implemented \citep{world_health_organization_international_2005}. 
Active surveillance aims to detect every case, not relying on a reporting structure and can have many approaches from sentinel sites or networks of sites that capture cases of a given condition, such as respiratory tract infections, within a catchment population \citep{murray_infectious_2017, melo-cristino_estudo_2006}. 
The application of environmental surveillance methods, performed prospectively to detect pathogens prior to the recording of clinical cases or to monitor their abundance in the environment to assess the potential risk of disease, has been proven as a viable alternative, particularly in wastewater \citep{andrews_environmental_2020, mcweeney_demonstration_1894, baker_combined_2011, larsen_tracking_2020}.

@@

\begin{figure*}[h!]
\centering
\includegraphics[width=\textwidth]{figures/introduction/Figure 3.pdf}
\caption{\textbf{Principles of current processing of bacterial pathogens based on whole genome sequencing.} Schematic representation of the workflow for processing samples for bacterial pathogens after the adoption of whole genome sequencing, with an expected timescale that could fit within a single day. The culture steps would be the same as currently used in a routine microbiology laboratory (see Figure \ref{fig:figure2}). Once a likely pathogen is ready for sequencing, \ac{DNA} is extracted, taking as little as 2 hours to prepare the \ac{DNA} for sequencing.
After sequencing, the main process for yielding information is computational. Automated sequence assembly algorithms are necessary for processing the raw sequence data, from which species, relationship to other isolates of the same species, antimicrobial resistance profile and virulence gene content can be assessed. All the results can also be used for outbreak detection and infectious diseases surveillance. Adapted from \cite{didelot_transforming_2012}}
\label{fig:figure3}
\end{figure*}

The paper describing the \ac{DNA}-sequencing method with chain-terminating inhibitors used in the sequencing of the first microbial genome \citep{sanger_dna_1977}, earned the late Frederick Sanger his share of the 1980 Nobel Prize in Chemistry alongside Walter Gilbert, was, in 2014, the top fourth in the number of citations with over 60000, highlighting its impact in the field of biological sciences, and by extension medicine \citep{van_noorden_top_2014}. 
Currently, this number has increased to over 84000 according to \ac{PMC}\footnote{\url{https://pubmed.ncbi.nlm.nih.gov/}}\footnote{\url{ https://www.ncbi.nlm.nih.gov/pmc/articles/PMC431765/}}. 
Since its emergence, reductions in cost, technical advances in sequencing technologies, and new computational developments have made genomic sequencing one of the most influential tools in biomedical research, yielding unprecedented insights into microbial evolution and diversity, and the complexity of the genetic variation in both commensal and pathogenic microbes. 
The emerging application of genomic technologies in the clinic to combat infectious diseases is transforming clinical diagnostics and the detection and surveillance of outbreaks.

@@

\begin{figure*}[h!]
\centering
\includegraphics[width=\textwidth]{figures/introduction/Figure 5.png}
\caption{\textbf{The three revolutions in sequencing technology that have transformed the landscape of bacterial genome sequencing.} The first-generation, also known as Sanger sequencers, is represented by the ABI Capillary Sequencer (Applied Biosystems). During the sequencing reaction, at each nucleotide incorporation event, a fluorescently labelled \ac{ddNTP} is incorporated, terminating the elongation of the \ac{DNA} molecule. The resulting electropherogram for sequencing reaction is below and is read from left to right. The second-generation, also known as high-throughput sequencers, is represented by MiSeq, a 4-channel sequencer, and NextSeq, a 2-channel sequencer (Illumina), both sequencing by synthesis instruments. For both instruments, the loaded flowcell is sequenced in massive parallel reactions, with each nucleotide incorporation emitting a light signal that is captured and later basecalled into a fastq file, with an indication of the confidence of the call, presented below. In a 4-channel instrument, each nucleotide has its own marker (A: yellow, T: green, C: red, G: blue) but in a 2-channel instrument, only 2 markers exist (A: green plus red, T: green, C: red, G: no marker). These instruments allow the sequencing of both ends of the \ac{DNA} fragment. Lastly, the third-generation, also known as long-read sequencers, is represented by the Pacific Bioscience BS sequencer and Oxford Nanopore MinION sequencer. In the first, immobilised polymerases in a \ac{SMRT} Cell incorporating nucleotides with identifying fluorescent labels. In the latter, a nanopore embedded in a solid-state membrane causes a change in an ionic current across the membrane each time a nucleotide is pushed through the pore. This difference in potential is then used for basecalling. Adapted from \cite{hagemann_overview_2015, loman_twenty_2015,goodwin_coming_2016, wang_nanopore_2021, metzker_sequencing_2010, xu_recent_2020}}
\label{fig:figure5}
\end{figure*}

@@

The release of the first truly high-throughput sequencing platform in the mid-2000s heralded a 50,000-fold drop in the cost of \ac{DNA} sequencing in comparison with the first-generation technologies and led to the denomination of the second generation as next-generation sequencing (NGS) \citep{goodwin_coming_2016}. 
This trend has continued over the next two decades of continued development and improvement, allied to the emergence of benchtop sequencing platforms with high-throughput sequencing data and turnaround times of days, making it a standard in any microbiology and public health laboratories \citep{loman_twenty_2015}. 
Second-generation sequencing methods can be grouped into two major categories: (1) sequencing by hybridisation and (2) sequencing by synthesis.

@@

The latter is defined by the use of terminator molecules that are similar to those used in the first-generation sequencing, preventing elongation of the DNA molecule, but unlike the first methods, it is reversible. 
To begin the process, a \ac{DNA} template is primed by a sequence that is complementary to an adapter region, which will initiate polymerase binding to this double-stranded \ac{DNA} region. 
During each cycle, a mixture of all four individually labelled and 3'-blocked \ac{dNTP}s are added. 
After incorporation of a single \ac{dNTP} into each elongating complementary strand, the unbound \ac{dNTP}s are removed and the surface is imaged to identify which \ac{dNTP} was incorporated at each cluster by optical capture. 
The fluorophore and blocking group can then be removed and a new cycle can begin \citep{goodwin_coming_2016}. 
The Illumina systems, which use this technology, account for the largest market share for sequencing instruments compared to other platforms\footnote{\url{https://www.forbes.com/companies/illumina/?sh=774358a91aa6}}, allowing paired-end sequencing and having the highest throughput (from 25 million reads for a MiSeq instrument to  1.2 billion reads for a NextSeq instrument\footnote{\url{https://www.illumina.com/systems/sequencing-platforms.html}}), with read lengths ranging from 45 to 300 bases in length with high accuracy, albeit with long running times (4 to 55 hours), rendering this technology a good choice for many sequencing applications where large read length is not required \citep{loman_performance_2012, gupta_chapter_2014, hagemann_overview_2015}.

@@

\ac{ONT} makes use of nanopores in small, portable single-molecule sequencing devices, capable of generating ultra-long sequences in real-time at a relatively low cost. 
Biological nanopores are embedded in solid-state membranes within disposable flow cells which, when a \ac{DNA} strand passes through the pore driven by a motor protein, each nucleotide causes a change in an ionic current across the membrane, which is later base called \citep{hoang_long-reads-based_2022, loman_twenty_2015}. 
This process is free from fluorescence labels and amplification requirements, and after one strand is processed, the pore is available to sequence the next available strand. 
Sequence quality and length depend on the loaded library but are usually much lower than the alternative counterparts, and its throughput is dependent on the number and lifespan of the nanopore within the flowcell but is still much lower than the alternatives. 
Despite this, its portability, fast advances, and continued improvement of the flowcells make this a fast-adopted technology for long-read sequencing.

@@

Currently, sequencing technologies are used in a case-by-case approach, with their adoption being much more present in a research setting than in a diagnostic one. 
Sequencing is mostly used after a diagnostic through the identification of the causative agent has already been performed. 
Although substantial advances have been made in reducing response time, most of the current systems do not yet generate enough data fast enough for a truly rapid response for it to be used in the clinical setting \citep{goodwin_coming_2016}. 
High-throughput \ac{DNA} sequencing has found additional new applications in drug discovery and in functional genomics with, for example, SNP-based analysis to identify new drug targets \citep{loman_twenty_2015}.

Although the second-generation \ac{DNA} sequencing methods have shed light on fundamental aspects of microbial ecology and function, they suffer from issues associated with short read length (see \ref{ssec:_intro_2nd_gen_seq}) and cannot reliably reconstruct long repeats because of uncertainties in mapping read, even when paired-end sequencing is used. 
Third-generation sequencing methods (see \ref{ssec:_intro_3rd_gen_seq} The third-generation of \ac{DNA} sequencing) have become increasingly used in microbiology, although their accuracy and low throughput make them difficult to implement in a clinical diagnostic setting.

@@

In the 2009 swine-origin Influenza A H1N1 pandemic, the first complete genome was publicly available on the 25 of April 2009 (GenBank accession number FJ966079), about a month after records of increased flu activity in Mexico and 10 days after the first confirmed cases in California, United States of America \citep{smith_origins_2009, novel_swine-origin_influenza_a_h1n1_virus_investigation_team_emergence_2009}. 
By the time the pandemic was declared, on 11 June 2009, \cite{smith_origins_2009} reported the origins and evolutionary genomics of the pandemic influenza A variant with a collection of 813 complete influenza genome sets, 17 of which belong to the new swine influenza viruses (GenBank accessions numbers GQ229259–GQ229378). 
The \ac{MERS} pandemic, declared as such in 2015 \citep{piret_pandemics_2021}, had its first publicly available sequence on 5 of July 2015 (GenBank accession number KT006149)\citep{lu_complete_2015}, with a sequence from a camel, thought to be an intermediate host for the virus, available as early as 7 of March 2016 (GenBank accession number KU740200) \citep{kandeil_complete_2016, al-shomrani_genomic_2020}.

@@

Molecular barcoding approaches can be combined with second-generation high-throughput sequencing to achieve unprecedented depths of coverage in microbial community profiling, being defined as metataxonomics. 
For profiling bacterial species, the most popular approach is 16S \ac{rRNA} gene sequencing, an ~1500 base pair gene that encodes catalytic RNA that is part of the 30S ribosomal subunit. 
Traditionally, the variable regions of the 16S \ac{rRNA} gene (V-regions) are targeted, or ranges thereof (V1-V2, V1-V3, V3-V4, V4, V4-V5, V6-V8, and V7-V9), and are specific to bacterial genus (96\%) and for some, even species (87.5\%), \citep{srinivasan_use_2015, abellan-schneyder_primer_2021}. 
Moreover, dedicated 16S databases that include near full-length sequences for a large number of strains and their taxonomic placements exist, such as RDP\footnote{\url{http://rdp.cme.msu.edu/}}, Greengenes\footnote{\url{https://greengenes.secondgenome.com/}}, silva\footnote{\url{https://www.arb-silva.de/}} and NCBI's 16S ribosomal RNA project\footnote{\url{https://www.ncbi.nlm.nih.gov/refseq/targetedloci/}} \citep{cole_ribosomal_2009, desantis_greengenes_2006, pruesse_silva_2007}. 
The sequence of an unknown strain can be compared with the sequences in these databases after very closely related sequences are grouped into \ac{OTUs},  an operational definition used to classify groups of closely related individuals. This allows the deduction of probable taxonomy, with the assumption that sequences of $>$95\% identity represent the same genus, whereas sequences of $>$97\% identity represent the same species \citep{schloss_introducing_2005}.

@@

While viruses are an integral part of the microbiota, no universal viral marker genes are available to perform such taxonomic assignments. 
Amplification of whole viral genomes is possible and, in 2015, RNA extracted from whole blood, serum, re-suspended swabs and urine, after targeted amplification of the whole viral genome, proved invaluable in the track of the Ebola virus disease epidemic in West Africa, responsible for >11 thousand deaths, allowing for the characterisation of the infectious agent the determination of its evolutionary rate, signatures of host adaptation, identification and monitoring of diagnostic targets and responses to vaccines and treatments \citep{quick_real-time_2016}.
As an alternative, broad scope viral targeted sequence capture (TSC) panels offer depletion of background nucleic acids and improve the recovery of viral reads by targeting coding sequences from a multitude of viral genera, such as VirCapSeq-VERT Capture Panel\footnote{\url{https://sequencing.roche.com/content/dam/rochesequence/worldwide/resources/brochure-vircapseq-vert-capture-panel-SEQ1000117.pdf}} but do not guarantee the full recovery of the viral genome, and can present biases towards certain genera \citep{schuele_assessment_2020, wylie_enhanced_2015}.

@@

Clinical applications of \ac{SMg} derive their roots from the use of microarrays (see \secref{ssec:_intro_current_standards}), which it was successfully applied in in-depth microbiome analysis of different sites in the human body, it was the emergence of the second-generation sequencing technology and its high throughput of genomic data at a competitive price that made sequencing of all genomic content, \ac{DNA} and/or RNA, in a clinical sample a viable possibility for diagnostics (see \secref{ssec:_intro_2nd_gen_seq} The second-generation of \ac{DNA} sequencing) \citep{miller_basic_2009, palmer_rapid_2006, chiu_clinical_2019}. The first reported case that demonstrated the utility of \ac{SMg} was in 2014 with the clinical diagnosis of neuroleptospirosis in a 14-year-old immunodeficient and critically ill boy with meningoencephalitis by Wilson et al \cite{wilson_actionable_2014}, prompting appropriate targeted antibiotic treatment and eventual recovery of the patient. In this case, traditional methods, including an invasive brain biopsy, failed to provide answers, until the shotgun sequencing of cerebrospinal fluid identified 475 of 3,063,784 sequence reads (0.016\%) corresponding to leptospira, for which clinical assays were negative due to its very low abundance. Ever since many other reports of successful application of \ac{SMg} in clinical metagenomics have been reported. but all in edge cases where traditional diagnostic methods have failed or as proof-of-concept \citep{couto_critical_2018, vijayvargiya_application_2019, sanabria_shotgun-metagenomics_2020, hirakata_application_2021}.

@@

As stated previously (see \secref{sec:_intro_genomics_approach} and \secref{ssec:_intro_metagenomics}), one of the biggest challenges when dealing with genomic, and in particular metagenomic, data is the lack of golden standards. This is also applicable to the bioinformatic analysis, required due to the amount of data produced by genomic sequencing technologies. This is currently one of the bottlenecks in the deployment of sequencing technology in clinical microbiology as there is no standard on how to deal with the increasing amount of data produced in a fit-for-purpose manner \citep{carrico_primer_2018}.
